Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2015-11-30 jss/article.tex
jss update
[skip ci] jss update

\subsection{ACSNMineR}
Currently, ACSN maps covers signaling pathways involved in DNA repair, cell cycle,
cell survival, cell death, epithelial-to-mesenchymal transition (EMT) and cell
motility. The information included in ACSN is organized in three hierarchical
levels: a global map, which is composed of five individual maps, themselves
composed of many functional modules covering individual pathways. Each map in
ACSN covers hundreds of molecular players, biochemical reactions and causal
relationships between the molecular players and cellular phenotypes. ACSN
represents a large-scale biochemical reaction network of 4,826 reactions
involving 2,371 proteins, and is continuously updated and expanded. 
We have included the three hierarchical levels in the \pkg{ACSNMineR} package,
in order to be able to calculate enrichments at all three levels.

>>>>

\subsection{ACSNMineR}
Currently, ACSN maps covers signaling pathways involved in DNA repair, cell
cycle, cell survival, cell death, epithelial-to-mesenchymal transition (EMT) and
cell motility. Each of these large-scale molecular map is decomposed in a number
of functional modules. The maps themselves are merged into a global ACSN map.
Thus the information included in ACSN is organized in three hierarchical levels:
a global map, five individual maps, and a total of 55 functional modules. Each
ACSN map covers hundreds of molecular players, biochemical reactions and causal
relationships between the molecular players and cellular phenotypes.  ACSN
represents a large-scale biochemical reaction network of 4,826 reactions
involving 2,371 proteins, and is continuously updated and expanded.  We have
included the three hierarchical levels in the \pkg{ACSNMineR} package, in order
to be able to calculate enrichments at all three levels. The calculations are
made by counting the number of occurences of gene symbols (HUGO gene names) from
a given list of genes of interest in all ACSN maps and modules. Table
\ref{tab:table1} is detailling the number of gene symbols contained in all the ACSN
maps.

\begin{table}[h!]
 \centering
  \caption{ACSN maps included in the \pkg{ACSNMineR} package. Map: map name, Total: total
  number of gene symbols (HUGO), Nb mod.: number of modules, Min: mimimum
  number of gene symbols in the modules, Max: maximum number of gene symbols in
  the modules, Mean: average number of gene sybols per module. N.B.: one gene
  symbol may be present in several modules of the map.}
  \label{tab:table1}
  \begin{tabular}{l|c|c|c|c|c}
    \hline
    Map & Total & Nb mod. & Min & Max & Mean\\
    \hline
  ACSN global & 2137 & 55 & 2 &625& 85\\
  Survival  &1065&46  &1 &434 &54\\
  Apoptosis & 668&44 & 1& 382& 33\\
  EMT \& Cell motility &620 &41  &1 &615 &43\\
  DNA repair &337&48  &1 &69  &19\\
  Cell cycle &119&45  &1 &27  &8\\
  \hline

  \end{tabular}

\end{table}

The statistical significance of the counts in the modules is assessed by using
either the Fisher exact test \citep{fisher1922interpretation,
fisher1934statistical} or the hypergeometric test, which are equivalent for this
purpose \cite{rivals2007enrichment}.

The current ACSN maps are included in the \pkg{ACSNMineR} package, as a list of character matrices.

\begin{verbatim}
> length(ACSN_maps)
[1] 6
> names(ACSN_maps)
[1] "Apoptosis"    "CellCycle"    "DNA_repair"   "EMT_motility" "ACSN_master" 
[6] "Survival"    
\end{verbatim}

For each matrix, rows represent a module, with the name of the module in the
first column, followed by a description of the module (optional), and then
followed by all the gene symbols of the module. The maps will be updated
according to every ACSN major release.

The main function of the \pkg{ACSNMineR} package is the \code{enrichment}
function, which is calculating over-representation or depletion of genes in the
ACSN maps and modules.
